The socialization of meritocracy and market justice preferences at school

### 2024-08-05 paper.tex
Actualización paper anonimizado en .tex y orden de repo

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdfauthor={Equipo EDUMER},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\author{}
\date{}

@@

\section{The socialization of meritocracy and market justice preferences
at
school}\label{the-socialization-of-meritocracy-and-market-justice-preferences-at-school}

\subsection{Introduction}\label{introduction}

Since its origins, educational institutions have been related to the
idea of social mobility and access to better opportunities. Therefore,
the consistent evidence of the high level of social reproduction at the
school level represents a threat to the promise of education and a
meritocratic system (\citeproc{ref-bourdieu_reproduction_1990}{Bourdieu
and Passeron 1990}). A large part of the research in this field at an
international level has addressed the extent to which the social origin
of students affects their academic results and their life opportunities
(\citeproc{ref-vonhippel_test_2019}{Von Hippel and Hamrock 2019}),
confirming that schools have severe difficulties in closing the gaps of
origin. Besides this socioeconomic perspective on school opportunities,
recent research has addressed to what extent inequalities in the school
context are also influencing students' perceptions, beliefs, and
attitudes: Are social inequalities perceived at the school context? Are
they rejected by the students, particularly those who are worst-off in
socioeconomic terms? Or, Is there evidence at the school level that
social inequalities are tolerated and even justified?
(\citeproc{ref-batruch_belief_2022}{Batruch et al. 2022};
\citeproc{ref-wiederkehr_belief_2015}{Wiederkehr et al. 2015}).

In the present paper, we deal with the justification of social
inequalities by eighth-grade students in Chile, a country characterized
by a highly stratified educational system. In particular, we focus on
market justice preferences (\citeproc{ref-lane_market_1986}{Lane 1986}),
which refer to the preferences for distributing public goods (as health
and education) based on criteria such as competition and payment
capacity. Although from a rational point of view it could be expected an
opposition to market justice by the underprivileged majority, we argue
that in a social environment characterized by a promotion of
meritocratic ideals - as the school system - would lead to the opposite:
a larger market justice preferences.

Given that the school environment has an important focus on performance,
achievement and acknowledgment, meritocracy has been one of the
principal concepts used for understanding and even for justifying
performance differences among students. Meritocracy is a distributive
system based on the belief that people should be rewarded and promoted
based on their abilities, knowledge, and achievements
(\citeproc{ref-young_rise_1958}{Young 1958}). It is often seen as a way
to create equal opportunities and fairness, as individuals can rise to
positions of power and influence based on their own merit rather than
their background or connections. However, some argue that meritocracy
can actually lead to tolerating or even justifying social inequalities,
as it can create a hierarchy where those who already have resources and
advantages are more likely to succeed. In this regard, a great deal of
academic research about meritocracy delves into the assessment of to
what extent rewards and privileges in society are related to merit,
emphasizing the so-called unfulfillable promise of meritocracy
(\citeproc{ref-mijs_stratified_2016}{Mijs 2016}). Complementing this
agenda, a second and emerging research area deals with subjective
aspects of meritocracy, such as perceptions and beliefs.

{[}definición de market justice, welfare marketization, (Lindh){]}

{[}contexto Chileno market justice, dar cifras/porcentajes de
privatización de salud, educación y pensiones; asociar al tema de la
calidad (lo público es peor){]}

The perception of meritocracy refers to how individuals view and
understand the concept of meritocracy in their own society
(\citeproc{ref-duru-bellat_who_2012}{Duru-Bellat and Tenret 2012};
\citeproc{ref-castillo_meritocracia_2019}{Castillo et al. 2019}). This
perception can vary greatly depending on individual experiences, social,
economic, and cultural background. Some people may see meritocracy as a
fair and just system that allows anyone to succeed based on their
abilities and hard work. In contrast, others may view it as a myth or a
cover for existing power dynamics and inequality, serving to maintain
and even reinforce inequality
(\citeproc{ref-lampert_meritocratic_2013}{Lampert 2013};
\citeproc{ref-mijs_paradox_2021}{Mijs 2021}). Based in this last
perspective, we argue that individuals with a higher perception of
meritocracy will show a larger justification of social inequalities, as
individual achievement would be seen as rewarded and social policies as
less necessary (\citeproc{ref-batruch_belief_2022}{Batruch et al.
2022}).

Most of the research that has related meritocratic beliefs to inequality
justification so far has only considered adults, leaving aside the study
of how beliefs in this field develop at student age as well as the
impact of the school context and the family as the main socialization
agencies. Regarding schools, the way in which they deal with unequal
conditions of origin has been linked to the \emph{hidden curriculum}
(\citeproc{ref-chafel_schooling_1997}{Chafel 1997}), whereby students
learn about distributive norms in society and mechanisms of
justification of social differences. Based on recent studies that relate
school meritocracy to the justification of economic inequalities in the
adult population (\citeproc{ref-batruch_belief_2022}{Batruch et al.
2022}; \citeproc{ref-wiederkehr_belief_2015}{Wiederkehr et al. 2015}),
the central hypothesis guiding this research is that school-age students
with a higher perception of meritocracy - both at school and at the
societal level - will show a larger justification of social
inequalities, as individual achievement would be seen as appropriately
rewarded and social mechanisms for correcting inequalities as less
necessary (\citeproc{ref-batruch_belief_2022}{Batruch et al. 2022}).

\textasciitilde{}

{[}introducir distinción sobre percepción de meritocracia a nivel
escolar y a nivel de la sociedad{]}

{[}resaltar 3 focos de innovación: market justice preferences at school
level, relación con meritocracia, y esquema de meritocracia en la
escuela y en la sociedad{]}

The present paper deals with the association between the perception of
meritocracy and the justification of social inequalities, with two main
focuses. Firstly, it assesses the justification of social inequalities
in social policy domains, such as health, pensions, and education. We
argue that individuals who perceive more meritocracy would be more
willing to justify better services in these domains for those with
higher incomes. Secondly, we focus on the student-age population as we
point out that it is possible to track down the origin of meritocratic
beliefs (and their consequences) to early socialization processes. To
this regard, we take into account the family and the school as two main
socialization agencies that play a significant role in the socialization
of cultural beliefs by transmitting cultural norms, values, and
expectations to young people.

\subsection{Justification of
inequality}\label{justification-of-inequality}

Research on social stratification beliefs, which explore individual
perceptions of who deserves what and why
(\citeproc{ref-kluegel_beliefs_1987}{Kluegel and Smith 1987}),
highlights that people's explanations and justifications of social
inequality are closely tied to their judgments of deservingness. The
influence of ideologies (\citeproc{ref-wegener_dominant_1995}{Wegener
and Liebig 1995}) and cultural schemas
(\citeproc{ref-homan_being_2017}{Homan, Valentino, and Weed 2017}) is
pivotal in shaping these explanations by offering symbolic
representations that frame societal structures and expectations. While
significant attention has been paid to wage inequality, income
distribution, and payment differentials in the literature {[}Castillo
(\citeproc{ref-castillo_legitimacy_2011}{2011}); Evans et al., 2010;
Jasso (\citeproc{ref-jasso_how_1999}{1999}) ; Shariff, Wiwad, and Aknin
(\citeproc{ref-shariff_income_2016}{2016}){]}, there has been less
examination of public beliefs about which life domains should be
governed by market relations and even less about children's acceptance
or rejection of these market principles. This oversight is notable given
the extensive encroachment of market logic into public goods, welfare
policy, and social services over the past five decades
(\citeproc{ref-centeno_arc_2012}{Centeno and Cohen 2012};
\citeproc{ref-harvey_breve_2015}{Harvey 2015}), affecting areas such as
pensions, health services, and education. The justification of social
inequality based on marketype criteria has been conceptualized as the
individuals' adherence to one specific justice evaluation, the market
justice, that is, affording legitimacy to the allocation of goods and
services based on prices and individuals' ability to pay
(\citeproc{ref-boltanski_new_2005}{Boltanski and Chiapello 2005};
\citeproc{ref-lane_market_1986}{Lane 1986};
\citeproc{ref-streeck_citizens_2012}{Streeck 2012}).

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
    behavsci,
    article,
    submit,
moreauthors
]{mdpi}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

% TODO: Add custom LaTeX header directives here
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{plainnat}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The socialization of meritocracy and market justice preferences at school},
  pdfauthor={Juan Carlos Castillo; Mauricio Salgado; Kevin Carrasco; Andreas Laffert},
  pdfkeywords={market
justice, Meritocracy, socialization, family, schools, Chile},
  pdfcreator={LaTeX via pandoc},
  unicode=true,
	bookmarksopen={true},
	pdffitwindow=true, 
	colorlinks=true, 
	linkcolor=bluecite, 
	citecolor=bluecite, 
	urlcolor=bluecite, 
	hyperfootnotes=true, 
	pdfstartview={FitH},
	pdfpagemode= UseNone}

\firstpage{1} 
\makeatletter 
\setcounter{page}{\@firstpage} 
\makeatother
\pubvolume{1}
\issuenum{1}
\articlenumber{0}
\pubyear{\the\year{}}
\copyrightyear{\the\year{}}
%\externaleditor{Academic Editor: Firstname Lastname}
\datereceived{ } 
\daterevised{ } % Comment out if no revised date
\dateaccepted{ } 
\datepublished{ } 
%\datecorrected{} % For corrected papers: "Corrected: XXX" date in the original paper.
%\dateretracted{} % For corrected papers: "Retracted: XXX" date in the original paper.
\hreflink{https://doi.org/} % If needed use \linebreak
%\doinum{}
%\pdfoutput=1 % Uncommented for upload to arXiv.org
%\CorrStatement{yes}  % For updates

\Title{ The socialization of meritocracy and market justice preferences
at school }

% MDPI internal command: Title for citation in the left column
\TitleCitation{ The socialization of meritocracy and market justice
preferences at school }

% Author Orchid ID: enter ID or remove command
%\newcommand{\orcidauthorA}{0000-0000-0000-000X} % Add \orcidA{} behind the author's name
%\newcommand{\orcidauthorB}{0000-0000-0000-000X} % Add \orcidB{} behind the author's name
\newcommand{\orcidauthorA}{0000-0003-1265-7854} 
\newcommand{\orcidauthorB}{0000-0002-8108-6154} 
\newcommand{\orcidauthorC}{0000-0002-9348-8361} 
\newcommand{\orcidauthorD}{0000-0002-9008-2454}

% Authors, for the paper (add full first names)
%\Author{Firstname Lastname $^{1,\dagger,\ddagger}$\orcidA{}, Firstname Lastname $^{2,\ddagger}$ and Firstname Lastname $^{2,}$*}

\Author{Juan Carlos Castillo$^{1}$\orcidA{}, 
Mauricio Salgado$^{2}$\orcidB{}, 
Kevin Carrasco$^{3}$\orcidC{}~and~Andreas Laffert$^{4}$\orcidD{}
}
%\longauthorlist{yes}

% MDPI internal command: Authors, for metadata in PDF
\AuthorNames{
  Juan Carlos Castillo,
  Mauricio Salgado,
  Kevin Carrasco,
  Andreas Laffert}

% MDPI internal command: Authors, for citation in the left column
%\AuthorCitation{Lastname, F.; Lastname, F.; Lastname, F.}
%\usepackage{xstring}
\newcommand\getfirst[1]{\getfirstaux#1\relax} \def\getfirstaux#1#2\relax{#1}
\newcommand{\authorformatted}{Castillo, \getfirst{Juan Carlos}.; Salgado, \getfirst{Mauricio}.; Carrasco, \getfirst{Kevin}.; Laffert, \getfirst{Andreas}.}
\AuthorCitation{\authorformatted}

% If this is a Chicago style journal: Lastname, Firstname, Firstname Lastname, and Firstname Lastname.

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address {%
$^{1}$ \quad Department of Sociology, Universidad de Chile\\
$^{2}$ \quad CEP, Centro de Estudios Públicos\\
$^{3}$ \quad COES, Center for Social Conflict and Cohesion Studies\\
$^{4}$ \quad Institute of Sociology, Universidad Católica de Chile\\
}

%\address{
%$^{1}$ \quad Affiliation 1; e-mail@e-mail.com\\
%$^{2}$ \quad Affiliation 2; e-mail@e-mail.com}

% Contact information of the corresponding author
%% see: https://quarto.org/docs/journals/authors.html
\corres{Correspondence: juancastillov@uchile.cl}

% Current address and/or shared authorship
%\firstnote{Current address: Affiliation.}  % Current address should not be the same as any items in the Affiliation section.
%\secondnote{These authors contributed equally to this work.}
% The commands \thirdnote{} till \eighthnote{} are available for further notes

%\simplesumm{} % Simple summary

%\conference{} % An extended version of a conference paper

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{ Previous research has shown that schools justify student
performance differences based on meritocratic ideals. However, the link
between school meritocracy and students' justification of inequalities
has received little attention. This paper examines how school and family
socialization influence students' beliefs about meritocracy and
market-based justifications for access to health, pensions, and
education in Chile. Using data from the First Study of Civic Education
in Chile, we employ a multilevel modeling strategy to analyze 5,047
students and parents nested within 231 schools. Our findings reveal that
a significant proportion of Chilean students agree with market justice
principles, at a higher rate than adults. Most students endorse
meritocratic views, particularly the reward of effort. Multilevel
estimations show strong associations between meritocratic beliefs and
market justice preferences: students who believe in the reward of effort
and talent are more likely to justify inequalities based on financial
capacity. At the school level, market justice preferences are higher in
high-status schools but lower in schools with better achievement.
Additionally, the conditional influence of meritocracy beliefs weakens
in schools with higher socioeconomic status and performance. These
results suggest that meritocratic beliefs and market justice preferences
begin early and are influenced by the school environment. }

% Keywords
\keyword{market
justice; Meritocracy; socialization; family; schools; Chile}

@@

\section{Introduction}\label{introduction}

Since its origins, educational institutions have been related to the
idea of social mobility and access to better opportunities. Despite
this, the consistent evidence of the high level of social reproduction
at the school level represents a threat to the promise of education and
a meritocratic system \citep{bourdieu_reproduction_1990}. A large part
of the research in this field at an international level has addressed
the extent to which the social origin of students affects their academic
results and their life opportunities \citep{vonhippel_test_2019},
confirming that schools have severe difficulties in closing the
socio-economic and cultural gaps of origin. Besides this socioeconomic
perspective on school opportunities, recent research has addressed to
what extent inequalities in the school context are also influencing
students' perceptions, beliefs, and attitudes: Are social inequalities
even perceived at the school context? Are they rejected by the students,
particularly those who are worst-off in socioeconomic terms? Or, Is
there evidence at the school level that social inequalities are
tolerated and even justified?
\citep{batruch_belief_2022, wiederkehr_belief_2015}.

Given that the school environment has an important focus on performance,
achievement and acknowledgment, meritocracy has been one of the
principal concepts used for understanding and even for justifying
performance differences among students. Meritocracy is a distributive
system based on the belief that people should be rewarded and promoted
based on their abilities, knowledge, and achievements
\citep{young_rise_1958}. It is often seen as a way to create equal
opportunities and fairness, as individuals can rise to positions of
power and influence based on their own merit rather than their
background or connections. However, some argue that meritocracy can
actually lead to tolerating or even justifying social inequalities, as
it can create a hierarchy where those who already have resources and
advantages are more likely to succeed
\citep{sandel_tyranny_2020, mcnamee_meritocracy_2004}. In this regard, a
great deal of academic research about meritocracy delves into the
assessment of to what extent rewards and privileges in society are
related to merit, emphasizing the so-called unfulfillable promise of
meritocracy \citep{mijs_stratified_2016}.

In the present paper, we address the role of the perception of
meritocracy on the justification of social inequalities by eighth-grade
students in Chile, a country characterized by a highly stratified
educational system. In particular, we focus on justifying inequalities
in health, education and pensions, which are traditional social policy
areas where access to better services could be justified based on
payment capacity, referred to as market justice
\citep{lane_market_1986, lindh_bringing_2023}. Most of the research in
this realm has only considered adults, leaving aside the study of how
beliefs in this field develop at student age as well as the impact of
the school context and the family as the main socialization agencies
\citep{batruch_belief_2022}. Regarding schools, the way in which they
deal with unequal conditions of origin has been linked to the
\emph{hidden curriculum} \citep{chafel_schooling_1997}, whereby students
learn about distributive norms in society and mechanisms of
justification of social differences.

Based on recent studies that relate school meritocracy to the
justification of economic inequalities in the adult population
\citep{batruch_belief_2022, wiederkehr_belief_2015}, the central
hypothesis guiding this research is that school-age students with a
higher perception of meritocracy - both at school and at the societal
level - will show a larger market justice preferences, as individual
achievement would be seen as appropriately rewarded and social
mechanisms for correcting inequalities as less necessary
\citep{batruch_belief_2022}. We focus on the student-age population as
we point out that it is possible to track down the origin of
meritocratic beliefs (and their consequences) to early socialization
processes. To this regard, we take into account the family and the
school as two main socialization agencies that play a significant role
in the socialization of cultural beliefs by transmitting cultural norms,
values, and expectations to young people.

The Chilean case is particularly intriguing for studying market justice
preferences. This country is characterized by acute and persistent
economic inequality, which stands out in Latin America and among OECD
countries \citep{flores_top_2020}. In Chile, the poorest 50\% captures
only 10\% of the total income and has negative wealth, while the richest
1\% receives almost 27\% of the income and holds 49.6\% of the wealth
\citep{chancel_world_2022}. Much of this inequality has been attributed
to the deep neoliberal reforms that institutionalized privatization and
commodification of various economic sectors. Such reforms where
introduced during the dictatorship (1973-1989) and expanded in democracy
through concessions, demand credits, and specific regulatory frameworks
\citep{ffrench-davis_reformas_2018}. This shift in economic policy has
allowed the unprecedented emergence of markets in areas related to
social reproduction, such as health, pensions, and education, with
provision and access managed by private entities and segmented by
individual payment capacity, heavily reliant on State subsidies
\citep{boccardo_30_2020}. In health, although the majority of the
population uses the public insurance system (78.9\%), 15.3\% are served
by private insurers \citep{observatoriosocial_estadisticas_2024}. The
pension system is based on individual capitalization, with mandatory
contributions managed by private administrators investing in the
financial market, currently involving 11 million contributors
\citep{superintendenciadepensiones_estadisticas_2024}. In education,
30.6\% of school enrollment is in public schools, 54.0\% in
state-subsidized (voucher) private schools, and 9.3\% in fully private
schools, generally attracting higher-income groups
\citep{ministeriodeeducacion_resumen_2023}.

\subsection{Justification of inequality and market
justice}\label{justification-of-inequality-and-market-justice}

The justification of social inequality based on market-type criteria has
been conceptualized as the individuals' adherence to the deservingness
of social goods and services (such as health, education, and pensions)
based on prices and individuals' ability to pay
\citep{boltanski_new_2005, lane_market_1986, streeck_citizens_2012}.
Research on social stratification beliefs, which explore individual
perceptions of who deserves what and why \citep{kluegel_beliefs_1987},
highlights that people's explanations and justifications of social
inequality are closely tied to their judgments of deservingness. The
influence of ideologies \citep{wegener_dominant_1995} and cultural
schemas \citep{homan_being_2017} is pivotal in shaping these
explanations by offering symbolic representations that frame societal
structures and expectations. While significant attention has been paid
to wage inequality, income distribution, and payment differentials in
the literature
\citep{castillo_legitimacy_2011, evans_family_2010, jasso_how_1999, shariff_income_2016},
there has been less examination of public beliefs about which life
domains should be governed by market relations
\citep{lindh_bringing_2023} and even less about children's acceptance or
rejection of these market principles. This oversight is notable given
the extensive encroachment of market logic into public goods, welfare
policy, and social services over the past five decades
\citep{centeno_arc_2012, harvey_breve_2015}, affecting areas such as
pensions, health services, and education.

There are substantial differences in funding and delivery methods in the
management of social services across nations
\citep{jensen_worlds_2008, stoy_worlds_2014}. Nordic countries, for
example, predominantly employ public agencies to produce and provide
social services, funding these through collective taxation and offering
them in kind to the majority of citizens. This system prioritizes social
justice, placing it above market mechanisms in accessing services. In
contrast, other countries rely more heavily on for-profit entities and
private funding, where service distribution depends mainly on individual
financial capacity to pay user fees, highlighting the influence of
market justice in service allocation. The trend toward marketization of
welfare services has been growing since the 1980s
\citep{salamon_marketization_1993}, and this shift is increasingly
evident even in countries where market solutions have traditionally had
a minor role in social policy \citep{sivesind_changing_2017}. The
expansion of marketization has been related to a larger justification of
market mechanisms, whereby societies with larger private spending on
services show larger market justice preferences
\citep{lindh_public_2015}.

Robert E. Lane proposed the underpinnings of the concept of market
justice, which he differentiated from political justice. For him, ``it
is the genius of the market to stimulate wants without at the same time
stimulating a sense of deserving more than one gets''
\citeyearpar[p.~384]{lane_market_1986}. Contrary to the evidence that
unequal distribution produces feelings of dissatisfaction, anger, and
resentment that might motivate forms of collective action
\citep{greitemeyer_subjective_2016, mishra_subjective_2015, smith_relative_2012, power_deprivationprotest_2018},
Lane pointed out that in market settings, social comparisons are more
likely to motivate increased effort rather than feelings of acute
injustice because individuals attribute outcomes to their actions. In
this sense, unequal levels of well-being would be, to some extent, a
function of their talents and efforts, instead of being based on
distributive principles that characterize welfare states, such as need
and equality (see \citep{wilson_role_2003}).

Despite high-income inequality and limited social mobility in Chile, and
in Latin America in general, there is a prevalent belief that
individuals are solely responsible for their economic outcomes, a view
that varies across the region
\citep{bucca_merit_2016, chong_mystery_2008, torche_intergenerational_2014, salgado_inequality_2023}.
The reliance on private welfare providers and widespread user fees
\citep{molyneux_neoliberal_2008} adds complexity to this context, as
reflected in surveys conducted by the Center for Public Studies (CEP).
According to this data, 35.9\% prefer private health insurance, and 63\%
would prefer private education
\citep{centrodeestudiospublicos_estudio_2024}. Yet, research on
children's justification in this area remains limited, highlighting a
significant gap in understanding how younger generations view
market-based access to welfare and whether these views are associated
with their meritocratic beliefs.

\subsection{Meritocratic perceptions and market
justice}\label{meritocratic-perceptions-and-market-justice}

The original definition of merit is a combination of effort and talent
\citep{young_rise_1958}, and a meritocracy is a distributive system
where merit is the main criterion for allocating valuable goods and
rewards. From a sociological perspective, meritocracy has been used in
research on social mobility to characterize societies with low mobility
that threaten the meritocratic ideal \citep{goldthorpe_myth_2003}. More
recently, sociology and social psychology research has attended to
subjective aspects related to the support for meritocratic principles in
different societies, such as beliefs in meritocracy
\citep{castillo_multidimensional_2023, mijs_unfulfillable_2016, mijs_paradox_2019}.
